### 2025-07-29 docs/report/Sections/benchmark.tex
Add report

\section{Benchmark Implementation}
\label{sec:bench}

This benchmark\footnote{\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}
is designed with a modular architecture, separating the \textit{view} and
\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
which allows for easy extension and integration of new template engines and
frameworks. It also includes a set of tests to ensure the correctness of
implementations and to validate the \textit{HTML} output. It includes two
different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class},
following the proposal of the
benchmarks~\cite{templatesbenchmark,springbenchmark}. The \texttt{Presentation}
class represents a presentation with a title, speaker name, and summary, while
the \texttt{Stock} class represents a stock with a name, URL, symbol, price,
change, and ratio.

@@

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes four main approaches to
PSSR\@:
\begin{itemize}
      \item \textbf{Reactive}: The template engine is used in a reactive context,
            where the HTML content rendered using the reactive programming model.
            An example of this approach is the Thymeleaf template engine
            when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
            non-blocking Spring \texttt{ViewResolver}.
      \item \textbf{Suspendable}: The template engine is used in a suspendable
            context, where the HTML content is rendered within the context of a suspending function.
            An example of this approach is the HtmlFlow template engine, which
            supports suspendable templates with the use of the \texttt{Flow} class from the
            Kotlin standard library.
      \item \textbf{Virtual}: The template engine is used in a non-blocking
            context, where the HTML content is rendered within the context of Virtual
            Threads. This method is used for the template engines that do not traditionally
            support non-blocking I/O, be it either because they use external DSLs and in
            consequence only support the blocking \texttt{Iterable} interface, or
            because they do not support the asynchronous rendering of HTML content.
      \item \textbf{Blocking}: The template engine is used in a blocking
            context, where the HTML content is rendered using the blocking interface of the
            \texttt{Observable} class, within the context of a OS-thread.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}. 
The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}},
this class is \textit{a controller method return value type for asynchronous
request processing where the application can write directly to the response
OutputStream without holding up the Servlet container thread.}

@@

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This limitation occurs because the response is only flushed to the client once
the content written to the \texttt{OutputStream} exceeds the configured output
buffer size, as explained in Section~\ref{sec:web-frameworks}. Consequently,
the client receives the complete response only after the entire HTML content
has been rendered and buffered, which negates the primary benefits of
PSSR—namely, reduced time-to-first-byte and progressive content delivery.
Furthermore, Spring MVC does not provide configuration options for the response
buffer size, preventing developers from reducing it to smaller values that
would enable more frequent flushing and achieve true progressive streaming of
HTML content. This architectural constraint makes Spring MVC unsuitable for
implementing effective PSSR compared to reactive frameworks like Spring
WebFlux.

@@

>>>>

\chapter{Benchmark Implementation}
\label{sec:bench}

This benchmark\footnote{\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}
is designed with a modular architecture, separating the \textit{view} and
\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
which allows for easy extension and integration of new template engines and
frameworks. It also includes a set of tests to ensure the correctness of the
implementations and to validate the HTML output.

\section{Data Model}

The benchmark includes two different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
shown in Listing \ref{lst:stock-class} and Listing \ref{lst:presentation-class}, 
following the proposal of the benchmarks~\cite{templatesbenchmark,springbenchmark}.
The \texttt{Presentation} class represents a presentation with a title, speaker
name, and summary, while the \texttt{Stock} class represents a stock with a
name, URL, symbol, price, change, and ratio.

@@

\section{Spring WebFlux}

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes four main approaches to
PSSR\@:
\begin{itemize}
  \item \textbf{Reactive}: The template engine is used in a reactive context,
        where the HTML content rendered using the reactive programming model.
        An example of this approach is the Thymeleaf template engine
        when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
        non-blocking Spring \texttt{ViewResolver}.
  \item \textbf{Virtual}: The template engine is used in a non-blocking
        context, where the HTML content is rendered within the context of Virtual
        Threads. This method is used for the template engines that do not traditionally
        support non-blocking I/O, be it either because they use external DSLs and in
        consequence only support the blocking \texttt{Iterable} interface, or
        because they do not support the asynchronous rendering of HTML content.
  \item \textbf{Suspendable}: The template engine is used in a suspendable
        context, where the HTML content is rendered within the context of a suspending function.
        An example of this approach is the HtmlFlow template engine, which
        supports suspendable templates with the use of the \texttt{Flow} class from the
        Kotlin standard library.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}.

\section{Spring MVC}

The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation,
this class is \textit{a controller method return value type for asynchronous
request processing where the application can write directly to the response
OutputStream without holding up the Servlet container thread.}

@@

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This is because the response is only sent once the content written to the
\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@.
As a result, the client receives the response only after the entire HTML
content is rendered, defeating the purpose of PSSR in this context.

@@

\section{Quarkus}

### 2025-07-29 docs/report/Sections/introduction.tex
Add report

\section{Introduction}

@@

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{scalability}. The two most dominant
approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
    Rendering (CSR)}. SSR generates HTML content on the server before sending it to
the client, resulting in a faster \textit{First Contentful Paint (FCP)}
~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). Nevertheless, SSR
can increase server load and reduce \textit{throughput}\footnote{The number of
    requests the server can handle per second (RPS)} since each request requires
additional processing before responding. In contrast, CSR shifts the rendering
workload to the browser—the server initially sends a minimal HTML document with
JavaScript, which dynamically loads the page content. While CSR reduces the
server’s burden, it can lead to a slower \textit{FCP}, as users must wait for
JavaScript execution before meaningful content appears.

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing progressive rendering as data becomes available, significantly reducing the
\textit{Time to First Byte (TTFB)} and improving perceived load times compared to traditional
SSR approaches~\cite{wise2024pssr}. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments.
Even so, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. Consequently, the client does
not need to execute JavaScript or make additional requests to retrieve page content.
The streaming nature of PSSR allows users to see content progressively as it becomes
available, rather than waiting for the complete page to be rendered server-side,
thus providing a more responsive user experience with measurably lower TTFB values~\cite{wise2024pssr}.

@@

With the introduction of \textit{Virtual
    Threads}~\footnote{\url{https://openjdk.org/jeps/444}} in Java 21, it is now
possible to execute blocking I/O operations in a scalable, lightweight manner.
This capability allows legacy template engines—often reliant on blocking
interfaces—to operate efficiently in high-concurrency, non-blocking
environments without requiring complex asynchronous programming models. Due to this, PSSR can now be implemented using familiar HTML templates, simplifying
development and improving maintainability.

We investigate the current landscape of non-blocking PSSR, focusing on two
primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
of which have been used to achieve asynchronous I/O in the Java ecosystem. As
an alternative, we investigate whether Java’s Virtual Threads can offer
comparable performance while preserving the simplicity of synchronous code.
Section 2 reviews the state-of-the-art in PSSR and template engine design.
Section 3 outlines the limitations of conventional engines in asynchronous
settings and presents our proposed approach. Section 4 details the benchmark
methodology, followed by the results and analysis in Section 5.
Section~6 compares our results with those from other studies.
Conclusions are presented in Section 7.

>>>>

\chapter{Introduction}

@@

\section{Motivation}

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{scalability}. The two most dominant
approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
    Rendering (CSR)}. SSR generates HTML content on the server before sending it to
the client, resulting in a faster \textit{First Contentful Paint (FCP)}
~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
can increase server load and reduce \textit{throughput}\footnote{The number of
    requests the server can handle per second (RPS)} since each request requires
additional processing before responding. In contrast, CSR shifts the rendering
workload to the browser—the server initially sends a minimal HTML document with
JavaScript, which dynamically loads the page content. While CSR reduces the
server’s burden, it can lead to a slower \textit{FCP}, as users must wait for
JavaScript execution before meaningful content appears.

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing progressive rendering as data becomes available. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments.
However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. As a result, the client does
not need to execute JavaScript or make additional requests to retrieve page content.

@@

With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
in Java 21, it is now possible to execute blocking I/O operations in a
scalable, lightweight manner. This capability allows legacy template
engines—often reliant on blocking interfaces—to operate efficiently in
high-concurrency, non-blocking environments without requiring complex
asynchronous programming models. As a result, PSSR
can now be implemented using familiar HTML templates, 
simplifying development and improving maintainability.

\section{Goals}

This work explores the current landscape of non-blocking PSSR, focusing on two
primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
of which have been used to achieve asynchronous I/O in the Java ecosystem. As an
alternative, we investigate whether Java’s Virtual Threads can offer comparable
performance while preserving the simplicity of synchronous code. Chapter 2
reviews the state-of-the-art in PSSR and template engine design. Chapter 3
outlines the limitations of conventional engines in asynchronous settings and
presents our proposed approach.

\section{Document Structure}

Chapter 4 details the benchmark methodology,
followed by the results and analysis in Chapter 5. Conclusions are presented in Chapter 6.

### 2025-07-29 docs/report/Sections/problem-statement.tex
Add report

\section{Problem Statement}

In this section, we examine the challenges of implementing \textit{Progressive
Server-Side Rendering} (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.
% Reactive streams~\cite{ReactiveStreams}, such as those based on
% \texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
% the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
% are abstractions that provide a non-blocking, asynchronous
% way to handle data streams.
% These abstractions support the incremental generation and transmission of HTML
% content as data becomes available, enabling \textit{progressive server-side
% rendering} (PSSR)~\cite{pssr2005}.
In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives.

Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
facilitate this by representing data as a sequence of asynchronous events.
For example, a reactive stream might emit a sequence of \texttt{Presentation}
objects, each representing a talk in a conference schedule.
As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in \autoref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}.
Note that the \texttt{await} builder receives an additional parameter, the
\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
proceed to render the next HTML element in the web
template~\cite{carvalho2023async}.
HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
similar to how the \emph{resume} function works in continuations and
coroutines~\cite{coroutines_continuations}.
In \autoref{lst:presentation-flow}, we show an
equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
Both examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\begin{center}
\begin{minipage}{0.50\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\textit{HtmlFlow reactive} presentation template in Koltin with an \texttt{Observable} model},
    label={lst:presentation-observable}
]
await { div, model, onCompletion ->
  model
    .doOnNext { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> div.raw(frag) }
      }
    .doOnComplete { onCompletion.finish() }
    .subscribe()
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.46\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={\textit{HtmlFlow suspend} presentation template in Kotlin with a \texttt{Flow} model},
]
suspending { model ->
  model
    .toFlowable()
    .asFlow()
    .collect { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}
\end{minipage}
\end{center}

By contrast, template engines that use \textit{external} DSLs—such as JStachio,
Thymeleaf, or Handlebars—typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{java.util.Iterable} or \texttt{java.util.stream.Stream}.
These interfaces require the entire data
model, to be materialized in memory before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as \texttt{Iterable} by blocking the thread until all items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

@@

These strengths make external DSLs appealing—even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
Emerging features in the Java ecosystem, particularly \textit{virtual threads}
introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. For this reason,
engines that rely on blocking interfaces—like those used in
external DSLs—can potentially achieve scalability levels that approach those of
non-blocking, asynchronous engines.

>>>>

\chapter{Problem Statement}

In this section, we examine the challenges of implementing \textit{Progressive
Server-Side Rendering} (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.
% Reactive streams~\cite{ReactiveStreams}, such as those based on
% \texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
% the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
% are abstractions that provide a non-blocking, asynchronous
% way to handle data streams.
% These abstractions support the incremental generation and transmission of HTML
% content as data becomes available, enabling \textit{progressive server-side
% rendering} (PSSR)~\cite{pssr2005}.

\section{Reactive and Suspendable Templates}

In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives. Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
facilitate this by representing data as a sequence of asynchronous events.
For example, a reactive stream might emit a sequence of \texttt{Presentation}
objects, each representing a talk in a conference schedule.
As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in Listing \ref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}.
Note that the \texttt{await} builder receives an additional parameter, the
\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
proceed to render the next HTML element in the web
template~\cite{carvalho2023async}.
HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
similar to how the \emph{resume} function works in continuations and
coroutines~\cite{coroutines_continuations}.
In Listing \ref{lst:presentation-flow}, we show an
equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
Both examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\begin{center}
\begin{minipage}{0.51\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\textit{HtmlFlow reactive} presentation template in Koltin with an \texttt{Observable} model},
    label={lst:presentation-observable}
]
await { div, model, onCompletion ->
  model
    .doOnNext { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> div.raw(frag) }
      }
    .doOnComplete { onCompletion.finish() }
    .subscribe()
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.47\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={\textit{HtmlFlow suspend} presentation template in Kotlin with a \texttt{Flow} model},
]
suspending { model ->
  model
    .toFlowable()
    .asFlow()
    .collect { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}
\end{minipage}
\end{center}

\section{External DSLs and Blocking Interfaces}

By contrast, template engines that use \textit{external} DSLs—such as JStachio,
Thymeleaf, or Handlebars—typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{java.util.Iterable} or \texttt{java.util.stream.Stream}.
These interfaces require the entire data
model to be materialized in memory before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as \texttt{Iterable} by blocking the thread until all items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O—especially under high concurrency. Listing \ref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

@@

These strengths make external DSLs appealing—even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
Emerging features in the Java ecosystem, particularly \textit{virtual threads}
introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. As a result,
engines that rely on blocking interfaces—like those used in
external DSLs—can potentially achieve scalability levels that approach those of
non-blocking, asynchronous engines, without requiring developers to adopt complex
asynchronous programming paradigms.
